---
title: "Analysis of the ToothGrowth dataset"
author: "Michal Siwek"
output: 
  pdf_document:
    fig_width: 6
    fig_height: 3
---

# Overview

This is part 2 of the Project Assignement for Statistical Inference course by JHU at Coursera.
It analyzes the ToothGrowth data in the R datasets package. In particular it shows:

* basic data explorations
* basic summary of data
* comparison of tooth growth by supp and dose using t-tests
* assumptions needed for performing the t-tests

# Exploratory Data Analysis and basic summaries of the data

Let's have a quick view of the data:
```{r}
head(ToothGrowth)
str(ToothGrowth)
summary(ToothGrowth)
hist(ToothGrowth$len)
```

Some exploratory plots:
```{r}
boxplot(tapply(ToothGrowth$len, ToothGrowth$dose, as.vector))
plot(ToothGrowth$supp, ToothGrowth$len)
```

Let's take a look at the exploratory plot offered for the ToothGrowth dataset in the help page of the dataset package.
```{r, fig.height=4}
require(graphics)
coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth,
       xlab = "ToothGrowth data: length vs dose, given type of supplement")
```

All these plots give a general suggestion that:
* the dose size has a positive impact on the tooth length
* orange juice gives higher tooth growth than ascorbid acid

# Comparing Tooth Growth by Supp and Dose

Comparison will be made using both **95% t confidence intervals** and **hypotheses tests at $\alpha=.05$**. I will show the calculations for supp in more detail, but I will skip most details for Dose, as the calculations run along the same lines in both cases.

## Comparing by Supp

Calculating auxiliary variables.
```{r}
g1 <- ToothGrowth[ToothGrowth$supp=='OJ',]$len
g2 <- ToothGrowth[ToothGrowth$supp=='VC',]$len
n1 <- length(g1); n2 <- length(g2)
s1 <- var(g1); s2 <- var(g2)
```

### Using Confidence Intervals

Let's first assume **paired observations**. The confidence interval is:
```{r}
diff <- g1-g2
mn <- mean(diff); s <- sd(diff)
mn + c(-1,1)*qt(.975,n1-1)*s/sqrt(n1)
```

The same can be achived using t.test function:
```{r}
t.test(diff)$conf.int
```

We see that **orange juce has a significantly more positive impact on the tooth growth**,
as the interval is entirely above 0.

Now let's assume **independent groups with equal variance**. The confidence interval is:
```{r}
mn <- mean(g1) - mean(g2)
pooled_var = ((n1-1)*s1 + (n2-1)*s2)/(n1+n2-2)
mn + c(-1,1)*qt(.975,n1+n2-2)*sqrt(pooled_var * (1/n1+1/n2))
```

The same can be achived using t.test function:
```{r}
t.test(g1, g2, paired=F, var.equal=T)$conf.int
```

We could also asssume **independent groups with unequal variance**:
```{r}
t.test(g1, g2, paired=F, var.equal=F)$conf.int
```

In both cases of independent group tests there is **no significant difference** between the groups.

### Using Hypotheses Tests

Let's reframe the problem in terms of hypotheses testing. Let's first assume **paired observations**. Test settings are:

* $H_0: diff = 0$
* $H-a: diff \neq 0$
* $\alpha = .05$

Critical value for two sided test:
```{r}
qt(0.975, n1-1)
```

Test statistic $TS = \frac{\bar{X} - \mu_0}{S / \sqrt{n}}$:
```{r}
(mn - 0) / (s/sqrt(n1))
```

so we **reject $H_0$** as the test statistic is greater than the critical value.

The same using t.test function:
```{r}
t.test(diff)
```

Now let's assume **independent groups with equal or unequal variance**. The test settings are:

* $H_0: mn1 = mn2$
* $H_a: mn1 \neq mn2$
* $\alpha = .05$

Assuming **equal variance**:
```{r}
t.test(g1, g2, paired=F, var.equal=T)
```

Assuming **unequal variance**:
```{r}
t.test(g1, g2, paired=F, var.equal=F)
```

I both cases of independent group tests we **can't reject $H_0$** as the p-values are above $.05$.

## Comparing by Dose

Comparing tooth length by dose runs along the same lines. Here I perform the tests only for paired observations. I compare each pair of doses, deducting sample means of bigger doses from those of smaller doses.

```{r}
doses <- combn(c(.5,1,2),2)
for(i in seq_len(ncol(doses))) {
    data <- subset(ToothGrowth, dose==doses[1,i])$len -
        subset(ToothGrowth, dose==doses[2,i])$len
    cat(">>>>> H0: dose of", doses[1,i], "vs. dose of", doses[2,i])
    print(t.test(data))}
```

We see that in each case that:

* the 95% t confidence intervals are entirely below $0$
* estimated means of differences are below $0$ with p-values below $.5$

so we conclude that **the bigger the dose of a supplement, the higher the tooth growth**.

# Conclusions and Assumptions

**Conclusions**:

* the dose size has a positive impact on the tooth length
* orange juice gives higher tooth growth than ascorbid acid

**Assumptions** required to hold for these conclusions:

* the sample of guinea pigs is a proper **random draw** (no confounding factors)
* the **observations are paired** - i.e. the same 10 guinea pigs were treated with every possible comination of supplement and dose
* **the order of pigs is preserved** in the data, i.e. the i-th row pig is the same pig under each treatment
* t-statistic assumes that **the underlying population have a normal distribution**, we should assume that it is close to normal (at least symmetric and mound shaped)
